---
title: "PUBH732 Week 6-9 <br> R Programming"
author: "Emma Wei Chen, Ph.D. <br> (wei.chen@ku.ac.ae)"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: simple
    footer: "KU-PUBH732-Spring2025-Emma"
editor: 
  markdown: 
    wrap: 72
---

# Lecture 2 <br>Data Management Using *tidyverse*

## Outline

### Topics:

-   Basic data management 
-   Cleaning data and heandling missing values 
-   Bias in data
-   Ethical considerations in data collection and analysis

## Outline

### Class Activities:

-   Run data management functions 
-   Cleaning a dataset by handling missing values and duplicates
-   Detect bias in data

## *tidyverse*

-   library("tidyverse")
-   Piping (%\>%)
    -  What is it?
    -  Why do you want to use it?

## Without Piping
Example: using mtcars dataset, The example filters cars with mpg > 20, groups them by cyl, and calculates the average hp for each group
```{r}
#| echo: true

## Example of Not using piping
filtered_summary <- summarise(group_by(filter(mtcars, mpg > 20), cyl), avg_hp = mean(hp, na.rm = TRUE))
```

## With Piping
```{r}
#| echo: true

## Example 1 of using piping
filtered_summary <- mtcars %>%
  filter(mpg > 20) %>%
  group_by(cyl) %>%
  summarise(avg_hp = mean(hp, na.rm = TRUE))

## Example 2 of using piping
filtered_summary <- mtcars %>% filter(mpg > 20) %>% group_by(cyl) %>% summarise(avg_hp = mean(hp, na.rm = TRUE))

```
## A good coding practice: using separate lines
-  Improved Readability
-  Simplifies Code Reviews
-  Easier Debugging
-  Enhances Maintainability


## Data Management Functions

-   mutate()
-   summarize()
-   group_by() and ungroup()

## Data Management Functions

-   filter()
-   select()
-   arrange()

## Data Management Functions

-   count()
-   rename()
-   row_number()
-   ifelse()


## Missing values

-   MCAR (Missing Completely at Random): The missingness is independent of any variable (randomly missing).
-   MAR (Missing at Random): The missingness depends on another observed variable (e.g., income is missing more frequently for younger individuals).
-   MNAR (Missing Not at Random): The missingness depends on the missing value itself (e.g., people with lower income tend not to report their income).


## End of the current lecture

### Any question/feedback?

## Class Activities:

-   Run data management functions 
-   Cleaning a dataset by handling missing values and duplicates
-   Detect bias in data

# Lecture 3 <br> Understanding the Nature of Data (Common Distributions: Normal & t-Distributions)

## Outline

### Topics:

Descriptive statistics: Mean, median, variance, standard deviation
Probability distributions: Normal distribution, t-distribution
Visualizing distributions using histograms and density plots Central
Limit Theorem and its implications

## Outline

### Class Activities:

Calculating summary statistics for a dataset Plotting histograms and
density curves for different datasets Comparing real-world examples of
normal and t-distributions

# End of the current lecture

## Any question/feedback?

# Lecture 4: Representing Data Using Statistical Tools and Standard Graphs

## Outline

### Topics:

Common statistical tools: Correlation, hypothesis testing basics Data
visualization: Histograms, boxplots, scatter plots Choosing the right
graph for different data types Best practices for effective data
representation

## Outline

### Class Activities:

Creating graphs (histograms, boxplots, scatter plots) using R/Python
Interpreting visualizations and discussing insights Identifying
misleading graphs and redesigning them for clarity


## Notes
Correlation results



## End of the current lecture

## Any question/feedback?
