---
title: "PUBH732 Weeks 6-9 <br> R Programming"
author: "Emma Wei Chen, Ph.D. <br> (wei.chen@ku.ac.ae)"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: simple
    footer: "KU-PUBH732-Spring2025-Emma"
    slide-number: true
    auto-stretch: true
    code-overflow: scroll

---

# Lecture 1 <br> Intorduction to R

## Outline

### Topics:

-   Why use R for data analysis?
-   Setting up the environment
-   Basic syntax and functionality of R
-   Loading datasets
-   Running simple scripts

## Outline

### Class Activities:

-   Running basic commands in R
-   Create a data frame
-   Loading datasets
-   Writing and executing a small script

## What tools did you use before?

## What tools did you use before?

### List of Other Statistical Tools

-   SPSS
-   MATLAB
-   STATA

### Why use R?

-   free of cost;
-   open access;
-   a large number of up-to-date packages;

## Textbook

Huynh, YW (2019) R for Graduate Students.
<https://bookdown.org/yih_huynh/Guide-to-R-Book/>

-   Other Materials

    Batra, Neale, et al. The Epidemiologist R Handbook. 2021.
    [![DOI](https://zenodo.org/badge/231610102.svg){alt="DOI"}](https://zenodo.org/badge/231610102.svg)
    <https://www.epirhandbook.com/en/> (Advanced with a focus on
    epidemiology)

## Setting up the environment

### Components

-   R
-   RStudio

### Installation

<https://posit.co/download/rstudio-desktop/>

## Setting up the environment

-   Creating a New Project (.Rproj)
-   Creating a Script (.R)
-   Layout of RStudio Interface
    -   Environment
    -   Script
    -   Console
    -   Package
    -   Working Directory

## Installing and loading packages

-   What are packages?
    -   A collection of free R tools that an R User wrote.
    -   Packages are free but must be installed to your computer first. After installation, packages must be loaded into your RStudio library each time RStudio is opened/launched.
        -   install.packages()
        -   library()
        
## Installing and loading packages

-   Different packages will provide different sets of tools, though some tools may have overlapping functions. In this case, the later loaded function will mask the earlier one.
    -   e.g. select() is in both *MASS* and *dplyr*. Use dplyr::select() to specify.
-   How do you know which package/function to use

## Useful packages

-   Useful packages:
    -   "*tidyverse*":
        -   This package is actually comprised of multiple packages:
            graphing (*ggplot2*) and user-friendly formatting (*dplyr*).
    -   "*writexl*": To produce Excel files from R data.
    -   "*readxl*": To load my Excel files into R.
    -   "*utils*": To read or write CSV files

## Basic syntax

-   R has all of the basic arithmetic operations available (+, -, /, \*, \^) and can function as a calculator. However, R is also a powerful tool for managing our data.
-   Frequently used operational symbols:
    -   ==, !=, \>, \>=, \<. \<=
    -   %in% (similar to == but much more useful)
    -   \|, &
-   Missing Value: NA

## Executing code

```{r}
#| echo: true

## Example 1
1 + 2 + 3 + 4

## Example 2
1 + 2 + 3 + # What happens if only select this line
  4

## Example 3
2 %in% c(1, 2, 3, 4)

## Example 4
1 + 2 + 3 + NA
```

## Executing code
-   Note the behavior of logic operations when NA is involved.

```{r}
#| echo: true

## Example 5
TRUE | FALSE

## Example 6
TRUE | NA

## Example 7
FALSE | NA
```

## Executing code
-   Note the behavior of logic operations when NA is involved.

```{r}
#| echo: true

## Example 8
TRUE & NA

## Example 9
FALSE & NA
```

## Using objects

-   Assigning values to objects (\<-)
-   What's the difference between \<- and = in R?
    -   \<- is used for general assignments to object
    -   = is used for parameters in functions

## Using objects

```{r}
#| echo: true

## Example for using <- 
x <- c(1, 2, 3, 4)
print(x)

y <- c(1, 2, 3, 4, NA)
## Example for using = 
mean(y, na.rm = FALSE)
mean(y, na.rm = TRUE)
```

## Using objects

-   Why you want to use objects
    -   save information in an object so you don't need to type them in again
    -   to apply functions on the object
-   Most commonly used object type in R: **data frame**
    -   data.frame() is the function to create a new data frame

## Data frame

```{r}
#| echo: true

## Example for create a data frame
df <- data.frame(
  Height = c(150, 60, 145, 187, 175), 
  Weight = c(55.5, 60.2, 62.3, 50.0, 48.1)
)

print(df)
```

## Data frame

-   Similar as a table.
-   Each column can have a different data type
    -   Commonly used data types: logical, integer, numeric, character,factor
```{r}
#| echo: true

## Example for logical data type
Smoker <- c(TRUE, FALSE, TRUE, FALSE, FALSE)

## Example for integer data type
ID <- c(1, 2, 3, 4, 5, 6, 7, 8)

## Example for numeric data type
df$Height

```

## Data frame

-   Note the difference between character and factor
```{r}
#| echo: true

## Add another column to df
df$City = c("Abu Dhabi", "Dubai", "Dubai", "Abu Dhabi", "Al Ain")
df$City_f = factor(df$City, levels = c("Abu Dhabi", "Dubai", "Al Ain", "Sharjah"))
summary(df)
```


## Data frame

-   Each row represents a single observation across all variables.
-   Columns have names (variable names), and rows can optionally have names.
-   You can access elements using `df[row, column]` notation.


## Data frame

```{r}
#| echo: true

# ## Add another column with less observations
# df$Gender = c("F", "M", "M")
# 
# ## Add another column with more observations
# df$Gender = c("F", "M", "M", "F", "F", "M")

## Add another column with the same number of observations
df$Gender = c("F", "M", "M", "F", "F")

## Add row names
rownames(df) <- c(1, 2, 3, 4, 5)
print(df)

```

## Datasets

-   Build-in Datasets
    -   data()
-   Datasets from R Packages. Below are PH&E related packages offers practice datasets:
    -   *pubh*
        [link](https://www.rdocumentation.org/packages/pubh/versions/2.0.0)
    -   *epi*
        [link](https://www.rdocumentation.org/packages/Epi/versions/2.59)
-   External data files
    -   read_xlsx() from *readxl* package
    -   read.csv() from *utils* package

## Writing a simple script

-   Header and Comment Organization
-   Naming (Note: R is case-sensitive!)
    -   Choose Descriptive and Concise Names

    -   Do Not Start with a Number or Symbol

    -   Avoid using punctuation other than a period (.), dash (-), or underscore (\_)

    -   No Spaces

    -   Consistency

    -   Avoid Names Already in Use
    
## Writing a simple script
-   A few naming styles
    -  Snake case (snake_case), e.g., group_mean
    -  Camel case (camelCase), e.g., groupMean
    -  Dot-separated (dot.case), e.g., group.mean
       -  This is not recommended because dot is used in some R functions, e.g. data.frame()
    -  Uppercase (UPPER_CASE), e.g., GROUP_MEAN


## Writing a simple script
-   My personal style
    -  I name key objects using Title Snake Case, e.g., Total_Count
    -  I name temporary objects using snake case, e.g., city_tbl 

## End of the current lecture

### Any questions/feedback?

## Class activities:

-   Practice basic commands in R
-   Create a data frame that include at least 3 different types of data: logical, integer, factor
    -   apply function summary() for your data frame to see what happens
-   Loading datasets
    -   try different ways to load datasets
-   Writing and executing a small script
    -   source()

## Before Next Class
-   Read "II Tidyverse" on the [R for Graduate Students](https://bookdown.org/yih_huynh/Guide-to-R-Book)